Demo arXiv template

### 2022-12-02 paper/paper.qmd
more on dimension reduction

@@

@@

Variable transformation, and scaling in the next section, are both pre-processing steps to create variables that fits assumptions for further computing. These include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
Variable transformation is a general notion of a functional transformation on the variable:

@@

@@

For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. All the scaling operations in green result in the same distribution as the original data.

@@

When constructing indexes from multivariate information, dimension reduction methods combine that information into a univariate series. In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):

@@

where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science. It can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.

*add another perspective that there are statistical methods, i.e. PCA, and also domain knowledge driven methods, i.e. d = P - PET in SPEI.*

The PCA is a special case of weighting, where individual variables are summed up in a linear combination with potential restrictions imposed on the weight coefficients. These coefficients may be simple mathematical constructions that give equal weight to each variable or can be derived with specialised knowledge by experts in the field. 
While suggested weights can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit specific tasks.

<!-- sometimes called feature extraction in the machine learning community  -->
<!-- With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.] -->

>>>>

*variables == indicators*

@@

*Effective drought index*

@@

The purpose of variable transformation is to create variables that fits assumptions for further computing. These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline. Variable transformation is a general notion of a functional transformation on the variable:

@@

<!-- A common assumption is normality in drought index.  -->

@@

For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.

@@

When indexes are constructed from multivariate information, dimension reduction methods combine that information into a univariate series. In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):

@@

where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science. It can be seen as a special case of weighting, where variables are summed up in a linear combination: $$x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$ with restrictions imposed on the weight coefficient: $\sum_{p=1}^P\lambda_p^2 = 1$. In other cases of weighting, the coefficients can be as simple as giving equal weight to each variables.

Some dimension reduction can also be formulated from domain-specific knowledge. This can be theories that describe the physics of the phenomenon being indexed or practical formulations used to extract useful features from the raw variables. For example, in the index SPEI, a difference series is calculated between precipitation and potential evapotranspiration (PET) and the validity of this formulation is backed up by climate water balance model [Thornthwaite, 1948], which describes [...]. *Add another example of remote sensing variables i.e. NDVI = (NIR - Red) / (NIR + Red)?*

<!-- Spectral satellite imageries often collected at different bands and it is common to combine values from multiple bands to compose new variables relevant to the index. The most common example of this is the Normalized Difference Vegetation Index (NDVI), which uses near-infrared (NIR) ($x_1(s;t)$) and red channel ($x_2(s;t)$): -->

<!-- \begin{equation} -->
<!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->
<!-- \end{equation} -->

<!-- which is used to evaluate the vegetation from near-infrared (NIR) and red channel. -->

While suggested weights and formulas can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit the specific tasks.

### 2022-12-02 paper/paper.tex
more on dimension reduction

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

@@

@@

For example, a z-score standardisation can be written in the above form
with \(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\), a min-max
standardisation uses \(\alpha = \min[x(s_i, t_j)]\) and
\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
pre-processing operations and uses color to differentiate whether the
operation is a variable transformation or a scaling step. All the
scaling operations in green result in the same distribution as the
original data.

@@

where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
variable dimension from \(P\) to \(P^*\). The most commonly used
dimension reduction technique is Principal Component Analysis (PCA),
also called Empirical Orthogonal Function (EOF) in earth science. It can
be written as
\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
where \(\lambda_p\) is the loading of the PC1, derived from maximising
the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).

\emph{add another perspective that there are statistical methods,
i.e.~PCA, and also domain knowledge driven methods, i.e.~d = P - PET in
SPEI.}

The PCA is a special case of weighting, where individual variables are
summed up in a linear combination with potential restrictions imposed on
the weight coefficients. These coefficients may be simple mathematical
constructions that give equal weight to each variable or can be derived
with specialised knowledge by experts in the field. While suggested
weights can indicate norms adored by practitioners, analysts should be
given the flexibility to experiment with different combinations when
constructing indexes. This could help understand index behavior from its
sensitivity to the variables and suggest alternative weights that better
suit specific tasks.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\emph{variables == indicators}

@@

\emph{Effective drought index}

@@

For example, a z-score standardisation can be written in the above form
with \(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\), a min-max
standardisation uses \(\alpha = \min[x(s_i, t_j)]\) and
\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
pre-processing operations and uses color to differentiate whether the
operation is a variable transformation or a scaling step. While both
variable transformation and scaling are pre-processing steps, the
scaling operations in green show the same distribution as the original
data.

@@

where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
variable dimension from \(P\) to \(P^*\). The most commonly used
dimension reduction technique is Principal Component Analysis (PCA),
also called Empirical Orthogonal Function (EOF) in earth science. It can
be seen as a special case of weighting, where variables are summed up in
a linear combination:
\[x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
with restrictions imposed on the weight coefficient:
\(\sum_{p=1}^P\lambda_p^2 = 1\). In other cases of weighting, the
coefficients can be as simple as giving equal weight to each variables.

Some dimension reduction can also be formulated from domain-specific
knowledge. This can be theories that describe the physics of the
phenomenon being indexed or practical formulations used to extract
useful features from the raw variables. For example, in the index SPEI,
a difference series is calculated between precipitation and potential
evapotranspiration (PET) and the validity of this formulation is backed
up by climate water balance model {[}Thornthwaite, 1948{]}, which
describes {[}\ldots{]}. \emph{Add another example of remote sensing
variables i.e.~NDVI = (NIR - Red) / (NIR + Red)?}

While suggested weights and formulas can indicate norms adored by
practitioners, analysts should be given the flexibility to experiment
with different combinations when constructing indexes. This could help
understand index behavior from its sensitivity to the variables and
suggest alternative weights that better suit the specific tasks.
